### 2022-05-05 doc/src/MSLevel.tex
Reorganize local and custom major interrupts
And weaken rules on future major interrupts.

The Advanced Interrupt Architecture assigns other major interrupt
identities in the range 16--63 to additional \emph{local interrupts}
that arise within or in close proximity to the hart, often for
reporting errors.
A mechanism is also defined that allows software to
selectively delegate both local and custom interrupts to the next lower
privilege level, or in some cases to inject entirely virtual interrupts
into a lower privilege level.

@@

%-----------------------------------------------------------------------
\section{Interrupt allocations and default priorities}
\label{sec:majorIntrs}

The base {\RISCV} Privileged Architecture designates major interrupt
numbers 16 and above as being for platform or custom use.
The Advanced Interrupt Architecture allocates 24 of these interrupt
numbers to standardize as local interrupts, and leaves the rest for
custom use.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|l|l|}
\hline
Default        &                         & \\
priority order & Major interrupt numbers & \multicolumn{1}{c|}{Description} \\
\hline
\hline
Highest & 63, 62, 31, 30, 61, 60, & Even numbers are standard local
                                     interrupts \\
        & 59, 58, 29, 28, 57, 56, & \ or reserved; odd numbers are
                                     designated \\
        & 55, 54, 27, 26, 53, 52, & \ for custom use \\
        & 51, 50, 25, 24, 49, 48  & \\
\cline{2-3}
        & 11, 3, 7   & Machine interrupts:  external, software, timer \\
        & \ 9, 1, 5  & Supervisor interrupts:  external, software, timer \\
        & 12         & Supervisor guest external interrupt \\
        & 10, 2, 6   & VS interrupts:  external, software, timer \\
        & 13         & Counter overflow interrupt \\
\cline{2-3}
        & 47, 46, 23, 22, 45, 44, & Even numbers are standard local
                                     interrupts \\
        & 43, 42, 21, 20, 41, 40, & \ or reserved; odd numbers are
                                     designated \\
        & 39, 38, 19, 18, 37, 36, & \ for custom use \\
Lowest  & 35, 34, 17, 16, 33, 32  & \\
\hline
\end{tabular}
\end{center}
\caption{%
The default priority order for major interrupts 0--63.
Interrupt~63 has the highest default priority, and interrupt~32 has the
lowest default priority.%
}
\label{tab:defaultIntrPrios}
\end{table*}

Table~\ref{tab:defaultIntrPrios} shows the allocation of major
interrupt numbers 0--63, listed in default priority order from highest
to lowest.
Note that, for interrupts with major identities in the range 16--31, a
higher interrupt number conveys higher default priority, and likewise
for major identities in the range 32--63.
These two groups are interleaved together in the complete order, and
the Privileged Architecture's standard interrupts, 0--15, are inserted
into the middle of the sequence.

\begin{commentary}
The default priority order is arranged so that interrupts 0--31 can
potentially be an adequate subset on their own for \mbox{32-bit}
{\RISCV} systems.
\end{commentary}

\begin{commentary}
If the {\RISCV} Privileged Architecture defines interrupt~0, the
Advanced Interrupt Architecture needs it to have a default priority
lower than certain external interrupts.
See Sections \ref{sec:mtopi} and~\ref{sec:stopi} on CSRs \z{mtopi} and
\z{stopi}.
\end{commentary}

Interrupt numbers 64 and above are all designated for custom use.
If a hart implements custom interrupts with these numbers, their
positions in the default priority order must be documented for the
hart.
Default priority order is standardized by the Advanced Interrupt
Architecture only for interrupts 0--63.

@@

%-----------------------------------------------------------------------
\section{Defined local interrupts}

The {\RISCV} Privileged Architecture currently defines one local
interrupt:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
13 & Counter overflow interrupt\ \ (from extension Sscofpmf) \\
\end{displayLinesTable}
The Advanced Interrupt Architecture defines these additional
local interrupts, listed in order of decreasing default
priority:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
30 & Reserved for standard reporting of bus or system errors \\
58 & Tentatively reserved for per-core high-power/over-temperature events \\
54 & Reserved for standard high-priority RAS events \\
\noalign{\bigskip}
38 & Reserved for standard low-priority RAS events \\
18 & Debug/trace interrupt \\
\end{displayLinesTable}
\emph{RAS} is an abbreviation for \emph{Reliability, Availability, and
Serviceability}.

\begin{commentary}
The Advanced Interrupt Architecture reserves local interrupt numbers
for use by other {\RISCV} standards.
Complete facilities for managing bus or system errors,
high-power/over-temperature events, RAS events, or debug/trace
interrupts are not defined here.
\end{commentary}

The Advanced Interrupt Architecture does not itself require that all
local interrupts it defines be implemented.
A {\RISCV} hart may implement any subset of them, or none of them.
For whichever ones are implemented, the corresponding bits in CSRs
\z{mip}/\z{miph} and \z{mie}/\z{mieh} must be writable, and the
corresponding bits in \z{mideleg}/\z{midelegh} (if those CSRs exist
because supervisor mode is implemented) may each either be writable or
be hardwired to zero.

Even when a defined local interrupt is implemented, it is not
necessarily the case that all events matching the description for that
interrupt actually cause the local interrupt to be raised.
For example, a single {\RISCV} system might report bus errors in
multiple ways, depending on where in the system the error is detected.
Some bus errors might signal the standard local interrupt (number~30),
while others are reported as external interrupts, routed through an APLIC
or IMSIC.

@@

When a hart implements one of the standard local interrupts defined by
the {\RISCV} Privileged Architecture (number 13, 14, or 15) or by the
Advanced Interrupt Architecture (an even interrupt number in the
range 16--62), the corresponding bit position is writable in both
\z{mip}/\z{miph} and \z{mie}/\z{mieh}.
An occurrence of the interrupt event causes the interrupt-pending bit
in \z{mip}/\z{miph} to be set to one.
This bit then remains set until cleared by software.

@@

When multiple interrupt causes are ready to trigger simultaneously, the
interrupt taken first is determined by priority order, which may be the
default order specified earlier, or may be a modified order configured
by software.

>>>>

The Advanced Interrupt Architecture reserves another 24 major
interrupt identities for additional \emph{local interrupts}
that arise within or in close proximity to the hart, often for
reporting errors.
A mechanism is also defined that allows software to
selectively delegate both local and custom interrupts to the next lower
privilege level, or in some cases to inject entirely virtual interrupts
into a lower privilege level.

@@

%-----------------------------------------------------------------------
\section{Defined major interrupts and default priorities}
\label{sec:majorIntrs}

Table~\ref{tab:majorIntrs} lists all the major interrupts
currently defined for {\RISCV} harts.
At this time, the only standard major interrupts are those
specified by the {\RISCV} Privileged Architecture.

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|l|l|}
\hline
Default        &                         & \\
priority order & Major interrupt numbers & \multicolumn{1}{c|}{Description} \\
\hline
\hline
Highest & 11, 3, 7   & Machine interrupts:  external, software, timer \\
        & \ 9, 1, 5  & Supervisor interrupts:  external, software, timer \\
        & 12         & Supervisor guest external interrupt \\
        & 10, 2, 6   & VS interrupts:  external, software, timer \\
Lowest  & 13         & Counter overflow interrupt \\
\hline
\end{tabular}
\end{center}
\caption{%
The standard major interrupt codes, listed in default priority order.
Currently, all standard codes are defined by
the {\RISCV} Privileged Architecture.%
}
\label{tab:majorIntrs}
\end{table*}

\begin{table*}[h!]
\begin{center}
\begin{tabular}{|c|l|}
\hline
                          & \\
Major interrupt numbers   & \hspace{10mm}Category \\
\hline
\hline
\begin{tabular}{@{}c@{}}
  \ 0--12 \\
  13--15 \\
\end{tabular} &
    \begin{tabular}{@{}l@{}}
      Not local interrupts \\
      Local interrupts \\
    \end{tabular}
      $\left\}\begin{tabular}{l@{}}
        Assigned by the \\
        \quad Privileged Architecture \\
      \end{tabular}\right.$ \\
\hline
16--23                    & Local interrupts\\
24--31                    & \em Designated for custom use \\
32--47                    & Local interrupts \\
\mbox{\ $\geq \mbox{48}$} & \em Designated for custom use \\
\hline
\end{tabular}
\end{center}
\caption{Categorization of current and future major interrupts.}
\label{tab:majorIntrCategories}
\end{table*}

Of the major interrupts controlled by the Privileged Architecture
(numbers 0--15), the Advanced Interrupt Architecture categorizes the
counter overflow interrupt (code~13) as a \emph{local interrupt}.
It is assumed furthermore that any future definitions for
reserved interrupt numbers 14 and 15 will also be local interrupts.
Major interrupt numbers 16--23 and 32--47 are additionally reserved for
standard local interrupts that other {\RISCV} extensions may define.
The remaining major interrupts allocated to the Privileged
Architecture, numbers 0--12, are categorized as not local interrupts.
Taken altogether, Table~\ref{tab:majorIntrCategories}
summarizes the Advanced Interrupt Architecture's categorization
of all major interrupt identities.

\begin{commentary}
For the standard local interrupts not defined by the
{\RISCV} Privileged Architecture (numbers 16--23 and 32--47),
the current plan is to assign default priorities
in the order shown in this table:
\begin{center}
\begin{tabular}{|c|l|l|}
\hline
Highest & 47, 23, 46, 45, 22, 44, & \\
        & 43, 21, 42, 41, 20, 40  & \\
\cline{2-3}
        & 11, 3, 7   & Machine interrupts:  external, software, timer \\
        & \ 9, 1, 5  & Supervisor interrupts:  external, software, timer \\
        & 12         & Supervisor guest external interrupt \\
        & 10, 2, 6   & VS interrupts:  external, software, timer \\
        & 13         & Counter overflow interrupt \\
\cline{2-3}
        & 39, 19, 38, 37, 18, 36, & \\
Lowest  & 35, 17, 34, 33, 16, 32  & \\
\hline
\end{tabular}
\end{center}
Among interrupts 16--23, a higher interrupt number conveys
higher default priority, and likewise for interrupts 32--47.
These two groups are interleaved together in the complete order,
and the Privileged Architecture's standard interrupts, 0--15,
are inserted into the middle of the sequence.
This proposed default priority order is arranged so that
interrupts 0--31 can potentially be an adequate subset on their
own for \mbox{32-bit} {\RISCV} systems.

In actuality, future {\RISCV} extensions may or may not stick to
this plan for the default priority order of interrupts they define.
\end{commentary}

\begin{commentary}
In addition to the existing major interrupts of
Table~\ref{tab:majorIntrs}, the following
local interrupts are tentatively proposed,
listed in order of decreasing default priority:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
23 & Bus or system error \\
45 & Per-core high-power or over-temperature event \\
43 & High-priority RAS event \\
\noalign{\smallskip}
35 & Low-priority RAS event \\
17 & Debug/trace interrupt \\
\end{displayLinesTable}
\noindent
These local interrupts are expected to be
specified by other {\RISCV} extensions.
Be aware, this list is not final and may change
as the relevant extensions are developed and ratified.
\emph{RAS} is an abbreviation for \emph{Reliability, Availability, and
Serviceability}.
\end{commentary}

\begin{commentary}
If a future version of the {\RISCV} Privileged Architecture
defines interrupt~0, the
Advanced Interrupt Architecture needs it to have a default priority
lower than certain external interrupts.
See Sections \ref{sec:mtopi} and~\ref{sec:stopi} on CSRs \z{mtopi} and
\z{stopi}.
\end{commentary}

Interrupt numbers 24--31 and 48 and higher
are all designated for custom use.
If a hart implements any custom interrupts, their
positions in the default priority order must be documented for the
hart.

@@

@@

For whichever standard local interrupts are
implemented, the corresponding bits in CSRs
\z{mip}/\z{miph} and \z{mie}/\z{mieh} must be writable, and the
corresponding bits in \z{mideleg}/\z{midelegh} (if those CSRs exist
because supervisor mode is implemented) must each either be writable or
be hardwired to zero.
An occurrence of a local interrupt event causes the interrupt-pending bit
in \z{mip}/\z{miph} to be set to one.
This bit then remains set until cleared by software.

@@

When multiple interrupt causes are ready to trigger simultaneously, the
interrupt taken first is determined by priority order, which may be the
default order specified in the previous section
(\ref{sec:majorIntrs}), or may be a modified order configured
by software.

### 2022-05-05 doc/src/VSLevel.tex
Reorganize local and custom major interrupts
And weaken rules on future major interrupts.

Instead of a physical VS-level \z{iprio} array, a separate hardware
mechanism is provided for configuring the priorities of a subset
of interrupts for VS level, using hypervisor CSRs \z{hviprio1} and
\z{hviprio2}, plus, for RV32, their high-half partners, \z{hviprio1h}
and \z{hviprio2h}.
The subset of major interrupt numbers whose priority may be configured
in hardware are these:
\begin{displayLinesTable}[r@{\quad}l]
 1 & Supervisor software interrupt \\
 5 & Supervisor timer interrupt \\
13 & (Reserved interrupt) \\
14 & \quad '' \\
15 & \quad '' \\
16 & \quad '' \\
18 & Debug/trace interrupt \\
20 & (Reserved interrupt) \\
22 & \quad '' \\
24 & \quad '' \\
26 & \quad '' \\
28 & \quad '' \\
30 & (Reserved for standard reporting of bus or system errors) \\
\end{displayLinesTable}
For interrupts directed to VS level, software-configurable priorities
are not supported in hardware for custom interrupts or for any
interrupt numbers higher than~30.

\begin{commentary}
If priority configurability is desired for custom interrupts, it can
be supported by custom CSRs in addition to what is defined here for
standard interrupts.
\end{commentary}

@@

RV32 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
\end{displayLinesTable}
RV32 \z{hviprio1h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Reserved for priority number for interrupt 8; reads as zero \\
bits 15:8  & Priority number for interrupt 13 \\
bits 23:16 & Priority number for interrupt 14 \\
bits 31:24 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV32 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 18, debug/trace interrupt \\
bits 23:16 & Priority number for interrupt 20 \\
bits 31:24 & Priority number for interrupt 22 \\
\end{displayLinesTable}
RV32 \z{hviprio2h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 24 \\
bits 15:8  & Priority number for interrupt 26 \\
bits 23:16 & Priority number for interrupt 28 \\
bits 31:24 & Priority number for interrupt 30 \\
\end{displayLinesTable}

@@

RV64 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
bits 39:32 & Reserved for priority number for interrupt 8; reads as zero \\
bits 47:40 & Priority number for interrupt 13 \\
bits 55:48 & Priority number for interrupt 14 \\
bits 63:56 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV64 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 18, debug/trace interrupt \\
bits 23:16 & Priority number for interrupt 20 \\
bits 31:24 & Priority number for interrupt 22 \\
bits 39:32 & Priority number for interrupt 24 \\
bits 47:40 & Priority number for interrupt 26 \\
bits 55:48 & Priority number for interrupt 28 \\
bits 63:56 & Priority number for interrupt 30 \\
\end{displayLinesTable}

@@

\begin{commentary}
For standard interrupt numbers in the range 13--63, and for
software interrupts (SSI), the corresponding interrupt-pending
bits in \z{vsip}/\z{vsiph} are defined as ``sticky,''
meaning a guest can clear them only by writing directly
to \z{sip}/\z{siph} (really \z{vsip}/\z{vsiph}).
Among the standard-defined interrupts, that leaves only
timer interrupts (STI), which can potentially be cleared
in \z{vsip} by writing a new value to \z{vstimecmp}.
\end{commentary}

>>>>

Instead of a physical VS-level \z{iprio} array, a separate hardware
mechanism is provided for configuring the priorities of a subset
of interrupts for VS level, using hypervisor CSRs \z{hviprio1} and
\z{hviprio2}, plus, for RV32, their high-half partners, \z{hviprio1h}
and \z{hviprio2h}.
The subset of major interrupt numbers whose priority may be configured
in hardware are these:
\begin{displayLinesTable}[c@{\quad}l]
\ 1    & Supervisor software interrupt \\
\ 5    & Supervisor timer interrupt \\
 13    & Counter overflow interrupt \\
14--23 & \em Reserved for standard local interrupts \\
\end{displayLinesTable}
For interrupts directed to VS level, software-configurable priorities
are not supported in hardware for standard local interrupts
in the range 32--48.

\begin{commentary}
For custom interrupts, priority configurability may be
supported in hardware by custom CSRs, expanding upon
\z{hviprio1} and \z{hviprio2} for standard interrupts.
\end{commentary}

@@

RV32 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & \em Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
\end{displayLinesTable}
RV32 \z{hviprio1h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 8; reads as zero \\
bits 15:8  & Priority number for interrupt 13, counter overflow interrupt \\
bits 23:16 & Priority number for interrupt 14 \\
bits 31:24 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV32 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 17 \\
bits 23:16 & Priority number for interrupt 18 \\
bits 31:24 & Priority number for interrupt 19 \\
\end{displayLinesTable}
RV32 \z{hviprio2h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 20 \\
bits 15:8  & Priority number for interrupt 21 \\
bits 23:16 & Priority number for interrupt 22 \\
bits 31:24 & Priority number for interrupt 23 \\
\end{displayLinesTable}

@@

RV64 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & \em Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
bits 39:32 & \em Reserved for priority number for interrupt 8; reads as zero \\
bits 47:40 & Priority number for interrupt 13, counter overflow interrupt \\
bits 55:48 & Priority number for interrupt 14 \\
bits 63:56 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV64 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 17 \\
bits 23:16 & Priority number for interrupt 18 \\
bits 31:24 & Priority number for interrupt 19 \\
bits 39:32 & Priority number for interrupt 20 \\
bits 47:40 & Priority number for interrupt 21 \\
bits 55:48 & Priority number for interrupt 22 \\
bits 63:56 & Priority number for interrupt 23 \\
\end{displayLinesTable}

@@

\begin{commentary}
For the standard local interrupts
(major identities 13--23 and 32--47), and for
software interrupts (SSI), the corresponding interrupt-pending
bits in \z{vsip}/\z{vsiph} are defined as ``sticky,''
meaning a guest can clear them only by writing directly
to \z{sip}/\z{siph} (really \z{vsip}/\z{vsiph}).
Among the standard-defined interrupts, that leaves only
timer interrupts (STI), which can potentially be cleared
in \z{vsip} by writing a new value to \z{vstimecmp}.
\end{commentary}
